---
title: "Writing custom knockoff statistics"
author: "Maike Ahrens, Sebastian Voss"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{custom-knockoff-statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Scope

The `knockoff.statistics()` function can be used with custom knockoff statistics
provided via the `statistic` argument. This allows 
for full flexibility with respect to the choice of modelling approach and/or 
feature importance measures, for instance to achieve consistency 
with existing workflows and downstream analyses.

Currently, `knockofftools` ships the following knockoff filters:

* Prognostic filters  
  * `stat_glmnet()`  
  * `stat_random_forest()`  
* Predictive filters
  * `stat_predictive_glmnet()`  
  * `stat_predictive_causal_forest()`  

This vignette outlines two levels of customization of the knockoff statistic 
computation:

1. Providing a custom filter function `stat_custom()` to be passed directly to 
`knockoff.statistics()` without further changes to the workflow. 
The worked example here covers the `{ranger}` implementation for a prognostic
filter (see [here](#lev1)). 
1. A version for the advanced user that wants to access intermediate results 
from the modelling and/or importance computing steps. To this end, one needs to 
conduct the individual steps wrapped in the package's work horse 
function `knockoff.statistics()` (see [here](#lev2)).


Start by loading the package and any additional dependencies
for the custom statistic.

```{r setup, message = FALSE}
library(knockofftools)
# dependencies for the custom statistic; adjust as needed
library(ggplot2)
library(purrr)
library(dplyr)
library(ranger)
library(stringr)
library(forcats)
# for Shapley value computation of tree-based models
library(treeshap)
# for analysis of survival outcomes
library(survival)

```


<!-- The knockoff filter requires a *feature statistic* $W_j$ for each predictor $j$. -->
<!-- while a near-zero or negative value suggests that $X_j$ and its knockoff copy -->
<!-- $\tilde{X}_j$ are exchangeable without meaningful loss of information. -->

## Level 1: Custom knockoff statistic {#lev1}

### The `stat_*()` interface {#stat-skeleton}

Let's start with the general structure of a function that can be
passed to `knockoff.statistics()` as a custom statistic, 
particularly its input arguments and the output it needs to return 
before going over its individual components in the next paragraphs.

Any function satisfying this interface can be passed to `knockoff.statistics()` 
via the `statistic` argument.

We are using [roxygen2](https://roxygen2.r-lib.org/)-style comments to
document the function.


```{r, eval = FALSE}
#' Custom knockoff feature statistic
#'
#' Computes a vector of feature statistics \eqn{W_j} that measure the
#' association between each predictor \eqn{X_j} and the response, adjusted
#' for its knockoff copy \eqn{\tilde{X}_j}. Positive, large values of `W_j` are
#' evidence that predictor `X_j` is associated with the response 
#' while a near-zero or negative value suggests that $X_j$ and its knockoff copy
#' $\tilde{X}_j$ are exchangeable without meaningful loss of information. 
#'
#' @param X `data.frame` of original predictors (numeric and/or factor columns).
#' @param X_k `data.frame` knockoff copy of `X`, same dimensions and column  
#' types as `X`. Consider the knockoff generation done at the point when  
#' `stat_custom()` is called (handled earlier in [knockoff.statistics()])
#' @param y response vector of length `nrow(X)` or `Surv` object 
#' (from package `survival`). numeric, binary `factor`, or `Surv` object.
#' @param type character string describing the outcome type. one of `"regression"` (the default), 
#' `"classification"`, or `"survival"`.
#' @param ... optional arguments forwarded from `knockoff.statistics()` to 
#' control modelling and importance computation.
#'
#' @return A `data.frame` with a single column `W` (one row per predictor) and
#' `rownames` equal to `colnames(X)`.
 
stat_custom <- function(X, X_k, y, type = "regression", ...) {
  
  # <input checks> 
  # <swapping for unbiased statistic>
  
  # <modelling and importance computation>
  
  # <return> 
}
```

## Worked example: `stat_ranger()`

For this vignette's example, a random forest is trained via 
(`ranger::ranger()`) and uses the built in option to compute 
permutation importance to derive importance scores for the knockoff filter. 
We build it in layers so each design decision is visible.
In comparison to the built-in `stat_random_forest()`, besides it 
fast implementation offers more commonly used options for splitrules, 
for instance.

#### Step 1: helper functions for modeling and scoring

The `scores_ranger()` function will be called inside the `stat_ranger()`
function. It takes care of fitting the random forest 
on the combined predictor matrix (original and knockoffs) and computing 
the importance values and it returns a named vector of permutation importance 
values, one per feature in the input matrix.
The function's scope mimics the existing score computation functions 
in the package, e.g. `scores_glmnet()`.


```{r, eval = !FALSE}

model_ranger <- function(X, y, type = "regression", ...) {
  # tackle potential errors specific to your modeling function of choice

  colnames(X) <- make.unique(colnames(X))  # ranger requires unique column names

  # Only forward arguments that ranger::ranger() actually accepts;
  # silently drop anything else to keep the interface safe when used
  # with knockoff.statistics() which may forward unrelated arguments.
  dots <- rlang::list2(...) %>%
    purrr::keep_at(rlang::fn_fmls_names(ranger::ranger))

  # handling of all three outcome types
  if (type == "survival") {
    rf_data    <- data.frame(time = y[, 1], status = y[, 2], X)
    rf_formula <- stats::as.formula("survival::Surv(time, status) ~ .")
  } else {
    rf_data    <- data.frame(y = y, X)
    rf_formula <- stats::as.formula("y ~ .")
  }
  
  do.call(
    ranger::ranger,
    list(
      formula    = rf_formula,
      data       = rf_data,
      importance = "permutation",
      num.trees  = 500,
      mtry       = ncol(X)   # all columns; X is cbind(X_orig, X_knock) here
    ) %>% 
      # allow updating defaults such as num.trees via ...
      purrr::list_assign(!!! dots)
  )
  
}

# the composed internal function for your custom knockoff statistic
scores_ranger <- function(X, y, type = "regression", ...) {
  
  # MODELLING
  rf_fit <- model_ranger(X = X, y = y, type = type, ...)
  
  # VARIABLE IMPORTANCE
  ranger::importance(rf_fit)
  
}
```

**Notes**: 

* Think of the function `scores_ranger()` as a generalization of your existing
workflow to obtain variable importances, independent of the knockoff 
application. The specifics for that will be handled outside of this function. 

* Handle all three `type` values if you intend the statistic to be
general-purpose. At minimum, throw an informative error for unsupported types
rather than letting the underlying model fail with a cryptic message.

* Similarly, handle all known potential issues related to your modelling and 
importance computation steps. This might be input checks related to naming
conventions for instance, but also related to the output say if your model 
did not converge or importances could not be computed.

* Consider splitting modelling and importance computation steps in separate 
functions to help with testing and maintenance. In this case, both are handled
directly by the `ranger::ranger()` call and we omit writing a separate function 
for the importance.  

* If shapley values are the preferred importance measure, 
the `treeshap` package can be used

```{r}
scores_ranger_shap <- function(X, y, type = "regression", ...) {
  
  # MODELLING as before   
  rf_fit <- model_ranger(X = X, y = y, type = type, importance = "none", ...)
  
  # VARIABLE IMPORTANCE
  treeshap::treeshap(
    treeshap::ranger.unify(
      rf_fit, 
      X %>% purrr::set_names(., make.unique(colnames(.))) %>% as.matrix()
    ), 
    X %>% purrr::set_names(., make.unique(colnames(.))) %>% as.matrix()
  )$shaps %>% 
    colMeans() %>% 
    abs()
}
```


* If you decide to write a custom statistic function for your engine
and/or importance measure of choice, you are most likely very familiar 
with their behavior and the potential pitfalls. 
These may range from formal requirements to the input structure to 
computational issues that may arise during the fitting or importance computation steps.
For instance, [ranger::ranger()] requires unique column names for the
input data and feature input to the custom statistic is a data set in which each
column name is duplicated (the generated knockoff data have the same column 
names as the original variables).
A lot harder to catch is the issue that in rare cases, 
NaN values may be returned for some out-of-bag predictions leading to 
missing values for variable importances. 
Make sure to catch potential errors and handle them gracefully, avoiding the 
computation to exit and render already completed knockoff iterations useless.
You may want to have a look into `purrr::safely()`, which may come in handy.
The details for making `stat_ranger()` bullet-proof will be skipped here, 
but handled in the implementation that will be 
part of an upcoming `{knockofftools}` release.


#### Step 2: `stat_ranger()` wraps helpers and eliminates bias

Fitting the forest naively on `cbind(X, X_k)` can introduce a directional
bias: the original columns always appear before the knockoff columns, so the
model may exploit column position rather than content. The standard fix is to
independently randomize which of $X_j$ and $\widetilde{X}_j$ occupies each
position before fitting, then correct the sign of $W_j$ afterwards.

Once `scores_ranger()` is in place, we wrap it in the `stat_ranger()`
function with the following structure (see [above](#stat-skeleton)): 

1. **Random swap**: for each $j$, exchange $X_j \leftrightarrow \widetilde{X}_j$
   with probability 0.5, independently across variables.
1. **Fit**: random forest on 
  $\text{cbind}(X_{\text{swap}},\,\widetilde{X}_{\text{swap}})$; 
  extract permutation importance for all $2p$
   columns.
1. **Difference**: $W_j = \text{imp}_j - \widetilde{\text{imp}}_j$.
1. **Sign correction**: multiply by $(1 - 2\, \cdot \mathbf{1}_{\text{swap}_j})$ to
   restore the correct orientation after the swap.


```{r, eval = !FALSE}
stat_ranger <- function(X, X_k, y, type = "regression", ...) {

  # input checks
  check_design(X); check_design(X_k)
  
  # 1. Randomly swap original and knockoff columns
  swap    <- as.logical(rbinom(ncol(X), size = 1, prob = 0.5))
  X_swap  <- X;   X_swap[, swap]  <- X_k[, swap]
  Xk_swap <- X_k; Xk_swap[, swap] <- X[, swap]
  
  # 2. Fit the forest on the combined matrix and extract importance
  imp <- scores_ranger_shap( # scores_ranger
    X    = cbind(X_swap, Xk_swap),
    y    = y,
    type = type,
    ...
  )

  # 3. Oftentimes, knockoff statistic is difference between 
  # original and knockoff importance
  p    <- ncol(X)
  orig <- seq_len(p)
  W    <- imp[orig] - imp[orig + p]

  # 4. Correct for swapped columns: flip the sign where columns were exchanged
  W <- W * (1 - 2 * swap)

  data.frame(W = W, row.names = colnames(X))
}
```




#### Step 3: end-to-end usage

Starting with a simulated dataset using the package's helper functions:

```{r, eval = !FALSE}
set.seed(1024)
X  <- generate_X(n = 200, p = 10, p_b = 0, cov_type = "cov_equi", rho = 0.3) 
lp <- generate_lp(X, p_nn = 3, a = 1) 
y  <- lp + rnorm(nrow(X))
```

Generate multiple knockoffs via `knockoff.statistics()` and
call `knockoff.statistics()` with our custom statistic `stat_ranger()`:


```{r}
W <- knockoff.statistics(
  y         = y,
  X         = X,
  type      = "regression",
  M         = 2,
  statistic = "stat_ranger"
)

S <- variable.selections(W, error.type = "pfer", level = 2)
S$stable.variables
```


```{r, fig.height = 5}
plot(S)
```

Additional `ranger::ranger()` arguments can be forwarded through `...`:

```{r, eval = FALSE}
# 1000 trees with a fixed internal seed
stat_ranger(X, X_k, y, type = "regression", num.trees = 1000, seed = 99)
```

### Survival and classification outcomes

`stat_ranger()` handles all three `type` values that `knockofftools` recognizes:

```{r, eval = FALSE}
# Classification
yb  <- factor(rbinom(nrow(X), 1, plogis(lp)))
X_k <- knockoff(X)
stat_ranger(X, X_k, yb, type = "classification")

# Survival (requires the survival package)
ys  <- simulWeib(N = nrow(X), lambda0 = 0.01, rho = 1, lp = lp)
X_k <- knockoff(X)
stat_ranger(X, X_k, ys, type = "survival")
```

### Making your statistic available to `knockoff.statistics()`

`knockoff.statistics()` resolves the `statistic` argument via `get()`, so your
function only needs to be on the search path at the time `knockoff.statistics()`
is called. There are two common patterns.

**Script or interactive session**: define the function before calling
`knockoff.statistics()`, either in the console or via `source()` when your 
functions live in a separate R file.

**R package**: for a hypothetical extension package `{knockoffranger}`
that contains `stat_ranger()`, simply make sure to 
export the function, install and attach the package, then pass
its name as a string:

```{r, eval = FALSE}
library(knockoffranger)   # exports stat_ranger()
W <- knockoff.statistics(y, X, statistic = "stat_ranger", M = 5)
```

Having the custom extension live in its own package has major benefits for 
sharing, testing and maintenance. 


## Level 2: Custom statistic with full flexibility in a manual workflow {#lev2}

Previously, we have shown how to implement a custom knockoff statistic, 
where the user is free to choose a combination of modelling engine and
importance metric. 
The output of `knockoff.statistics(statistic = "stat_ranger")`
however is still limited to the most streamlined version, which is the 
importance table of each variable per iteration. 

Here, we present a more granular manual workflow that allows 
for the extraction of additional information from the workflow. 
This is primarily of interest to the users who are more advanced in 
the machine learning field, and are keen to better understand for example how
the introduction of the knockoff variables affects their models, 
for example in terms of performance. 
Another simple extraction of interest might be the $2p$ vector of individual
importances for original and knockoff variables before their 
aggregation.

In the following, we'll extract this information alongside the
knockoff statistics $W$. 


```{r, fig.height=5}
# simulate original dataset
set.seed(1)
X   <- generate_X(n = 100, p = 10, p_b = 0, cov_type = "cov_equi", rho = 0.3)
p   <- ncol(X)
lp  <- generate_lp(X, p_nn = 3, a = 1)
y   <- lp + rnorm(nrow(X))

# generate knockoff data sets (M iterations)
M    <- 5
X_ks <- purrr::map(
  1:M, 
  ~ {set.seed(.x); knockoff(X)}
)

# draw swaps for columns of X and Xk
swaps <- purrr::map(
  1:M, 
  ~as.logical(rbinom(p, 1, 0.5))
)

# swapping helper
swap_it <- function(X, X_k, swap){ # , rename_ko = TRUE
  
  # check 
  # dim(X) == dim(X_k)
  # swap logical of length ncol(X)
  
  # generate modified versions of original and knockoff matrix
  X_swap  <- X;  
  X_swap[,swap] <- X_k[,swap]
  names(X_swap)[swap] <- paste0(names(X)[swap],  "_ko")
  
  Xk_swap <- X_k; 
  Xk_swap[,swap] <- X[,swap]
  Xk_swap <- Xk_swap
  names(Xk_swap)[!swap] <- paste0(names(X)[!swap],  "_ko")

  list(
    data = cbind(X_swap, Xk_swap), 
    swap = swap
  )
  
}


# initiate result dataframe for knockoff statistics from each iteration m = 1, ..., M
# fully in line with what knockoff.statistics() returns
# for compatibility with downstream usage, e.g. variable.selections(W)
W <- matrix(
  NA, 
  nrow = p, ncol = M, 
  dimnames = list(colnames(X), paste0("W", 1:M))
) %>% 
  as.data.frame()

# the manual workflow allows for the extraction of additional information, 
# the user has no access to in the standard workflow
# initiate list to store this information for each iteration
custom_info <- list()
  
for (i in 1:M) {
  # i <- 1
  
  swap_result <- swap_it(X, X_ks[[i]], swaps[[i]]) # , rename_ko = TRUE
  X_imp <- swap_result$data
  
  # modelling
  # using ranger engine like in simple customization example
  custom_model <- ranger::ranger(
    x = X_imp, 
    y = y, 
    num.trees  = 100, # not recommended for practical application
    importance = "permutation"
  )
  
  # variable importances
  # example using {treeshap} package for computation of Shapley values
  # individual contributions are averaged to obtain per variable importance
  custom_scores <- treeshap::treeshap(
    treeshap::ranger.unify(custom_model, as.matrix(X_imp)), 
    as.matrix(X_imp), 
    verbose = FALSE 
  )$shaps %>% 
  colMeans()
  # custom_scores
  
  # Compute custom statistic W 
  # difference in absolute mean Shapley values
  orig <- 1:p
  custom_importance <- (abs(custom_scores[orig]) - abs(custom_scores[orig + p])) *
    # correct for swapping of columns of X and Xk
    (1 - 2*swap_result$swap) %>% 
    rlang::set_names(names(X))

  W[,i] <- custom_importance
  
  custom_info[[i]] <- list(
    importance = list(
      permutation  = custom_model$variable.importance %>% 
        magrittr::set_attr("mode", custom_model$importance.mode),
      shapley = custom_scores
    ) %>% 
      as.data.frame() %>% 
      # optional cosmetics
      {
        if (require(naturalsort)) {
          {.[naturalsort::naturalsort(rownames(.)),]} %>% 
            {.[c(stringr::str_subset(rownames(.), "\\d$"), stringr::str_subset(rownames(.), "ko$")),]}
        } else {.} 
      }, 
    pred_error = custom_model$prediction.error,
    pred_oob   = custom_model$predictions
  )
}

W 

# the object W contains custom statistics and can be passed to downstream methods
S <- variable.selections(W, error.type = "pfer", level = 3)

plot(S)
```


```{r, fig.height = 5, fig.width = 5.25}
# summary of prediction errors across iterations
custom_info %>% purrr::map_dbl("pred_error") %>% summary()

# visual comparison of importance metrics per variable in a facets per metric   in a boxplot
custom_info %>% 
  purrr::map("importance") %>% 
  purrr::map(~ tidyr::pivot_longer(.x %>% tibble::rownames_to_column("feat"), cols = -feat, names_to = "metric", values_to = "importance") ) %>% 
  list_rbind(names_to = "iter") %>% 
  mutate(
    type = ifelse(feat %>% stringr::str_detect("ko$"), "knockoff", "original") %>% factor(levels = c("original", "knockoff")), 
    variable = stringr::str_remove(feat, "_ko$") %>% forcats::fct_relevel(colnames(X) %>% rev())
  ) %>% 
  
  ggplot(aes(x = importance, y = variable, fill = type)) +
  geom_vline(mapping = aes(xintercept = 0), linetype = "dashed", color = "grey") +
  geom_boxplot(position = position_dodge2()) +
  facet_wrap(~ metric, scales = "free_x") +
  theme_bw() +
  theme(legend.position = "top") +
  scale_fill_manual(values = c("original" = "steelblue", "knockoff" = "salmon")) + 
  labs(title = paste0("Comparison of importance metrics (", M, " iterations)"))
```

